---
title: "BactDate_example"
author: "Zoe A. Dyson, zoe.dyson(at)lshtm.ac.uk"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  prettydoc::html_pretty:
    theme: cayman
    number_sections: TRUE
    fig_width: 8
    fig_height: 8
---

## Setup knitr
```{r setup}
knitr::opts_chunk$set(echo = TRUE)
```

## Table of contents
- [Root-to-tip regression](#roottotip)  
- [BactDating MCMC model selection](#bactdatingmodelselection)
- [BactDating MCMC analysis plots](#bactdatingmcmcplots)
- [BactDating MCMC temporal signal](#temporaltest)
- [Output dated tree with annotations](#outputtree)

## Load packages & track versions
```{r load_packages}
# R version
R.Version()$version.string

# Reporting
library(knitr)
packageVersion("knitr")

# Wrangling
library(tidyverse)
packageVersion("tidyverse")

library(lubridate)
packageVersion("lubridate")

library(coda)
packageVersion("coda")

# Phylo
library(BactDating)
packageVersion("BactDating")

library(ape)
packageVersion("ape")

library(ggtree)
packageVersion("ggtree")

library(treeio)
packageVersion("treeio")

```

## Setup R environment
```{r setup_r_env}

# Set seed
set.seed(3551)

# Setup `not in` operator
`%notin%` <- Negate(`%in%`)

# Setup number chain length/iterations
chain_length <- 1e6

# Set working directory 
setwd("~/Dropbox/example_file_path")

```

## Import data
```{r import_data}

# Import post-gubbins tree & root on outgroup
tree <- loadGubbins("GUBBINS")

# Import metadata 
meta <- read_csv("metadata.csv")

```

## Format data for BactDating analyses
```{r wrangling}

# Get ingroup/outgroup sequences
ingroup_sequences <- meta %>%
  filter(LANE_ID %in% tree$tip.label) %>%
  select(LANE_ID) %>%
  pull()

outgroup_sequences <- tree$tip.label[tree$tip.label %notin% ingroup_sequences] 

# Remove outgroups from tree
for (tip in 1:length(outgroup_sequences)) {
  tree <- drop.tip.useRec(tree, outgroup_sequences[tip])
}

# Order dates vector to match order of tips
dates <- meta %>%
  mutate(collection_date = as.Date(date_enrolled,"%d/%m/%Y")) %>%
  mutate(decimal_years = decimal_date(collection_date)) %>%
  select(LANE_ID, decimal_years) %>%
  slice(match(tree$tip.label, LANE_ID)) %>%
  deframe() 

# Plot tree
plot(tree)
ape::axisPhylo(backward=F)

```

 
## Root-to-tip regression analysis for temporal signal {#roottotip}
```{r root_to_tip_regression}

# Root to tip regression
roottotip(tree,dates) 

```


## MCMC BactDating model selection {#bactdatingmodelselection}
```{r model_selection}

# Strict clock
bactdating_strict_mcmc <- bactdate(tree,dates,
                                   model='mixedgamma',
                                   nbIts=chain_length,
                                   useRec = T,
                                   updateRoot = F,
                                   showProgress = F)
bactdating_strict_mcmc
plot(bactdating_strict_mcmc,type="trace")
try(plot(bactdating_strict_mcmc, type="scatter"))
bactdating_strict_mcmc$pstrict

# Relaxed clock
bactdating_relaxed_mcmc <- bactdate(tree,dates,
                                    model='relaxedgamma',
                                    nbIts=chain_length,
                                    useRec = T,
                                    updateRoot = F,
                                    showProgress = F)
bactdating_relaxed_mcmc
plot(bactdating_relaxed_mcmc,type='trace')
try(plot(bactdating_relaxed_mcmc, type="scatter"))
bactdating_relaxed_mcmc$prelaxed

# Check if ESS sufficient for model comparison
effectiveSize(as.mcmc.resBactDating(bactdating_strict_mcmc))
effectiveSize(as.mcmc.resBactDating(bactdating_relaxed_mcmc))

# Compare models
modelcompare(bactdating_strict_mcmc,bactdating_relaxed_mcmc)

```


## MCMC select best fit model & plot {#bactdatingmcmcplots}
```{r mcmc_plots}

# Select best fitting model 
bactdating_mcmc <- bactdating_relaxed_mcmc 

# Plot tree with CI
plot(bactdating_mcmc,'treeCI',show.tip.label = F)

# Plot most likely root
plot(bactdating_mcmc,'treeRoot',show.tip.label=F)
```

## Test for temporal signal {#temporaltest}
```{r temporal_test}

# Force dates to be equal 
bactdating_mcmc_testdates <- bactdate(tree,
                                      rep(2015,length(dates)),
                                      nbIts=chain_length,
                                      model='mixedgamma',
                                      useRec = T,
                                      updateRoot = F,
                                      showProgress = F)

# Compare models
modelcompare(bactdating_mcmc,bactdating_mcmc_testdates)

# Randomise tip dates
rando_dates <- sample(dates)
names(rando_dates) <- names(dates)

bactdating_mcmc_randodates <- bactdate(tree,
                                      rando_dates,
                                      nbIts=chain_length,
                                      model='mixedgamma',
                                      useRec = T,
                                      updateRoot = F,
                                      showProgress = F)
# Compare models
modelcompare(bactdating_mcmc,bactdating_mcmc_randodates)

```


## Output beast-ish tree {#outputtree}
```{r output_dated_tree}

# Make beast-ish tree file
beast_tree <- as.treedata.resBactDating(bactdating_mcmc)
beast_tree <- methods::new('treedata',phylo=beast_tree[[1]],data=dplyr::tbl_df(as.data.frame(beast_tree[[2]])))

# Plot tree for sanity check
ggtree(beast_tree, mrsd=date_decimal(max(dates))) + 
  geom_range(range='length_0.95_HPD', color='red', alpha=.6, size=2) +
  geom_text(aes(label=node)) + 
  theme_tree2()

# Output tree file
write.beast(beast_tree, "BactDating_tree.tree")
```

